---
title: 'R Programming - Assignment #3'
author: "Myron Keith Gibert Jr"
date: "January 21, 2020"
output: pdf_document
header-includes:
- \usepackage{hyperref}
- \hypersetup{
    colorlinks=true,
    linkcolor=black,
    filecolor=magenta,      
    urlcolor=cyan,
    pdftitle={R Programming - Assignment 3},
    pdfpagemode=FullScreen,
    }
toc: TRUE
---

\thispagestyle{empty}
\clearpage
\pagenumbering{arabic} 
\pagebreak

```{r tinytex, include=FALSE}
## INSTALLING LATEX FOR RMARKDOWN

#RMarkdown requires LaTex to create pdf documents as an output. More information can be found [here](https://bookdown.org/yihui/rmarkdown/pdf-document.html). Alternatively, output can be set to "html_document" or "word_document". End users may install LaTex by setting the "wanttinytex" variable to FALSE and running the following script:

#Install tinytex to let RMarkdown create a pdf document? Default: wanttinytex <- FALSE
wanttinytex <- FALSE

if(wanttinytex == TRUE){

if(tinytex:::is_tinytex()==FALSE && wanttinytex == TRUE){
  if (!require("tinytex")) install.packages("tinytex")
  tinytex::install_tinytex()
  tinytex:::install_prebuilt()}

library("tinytex")
#tlmgr_install("fvextra")

}

```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)

if (!require("tidyverse")) install.packages("tidyverse")
library("tidyverse")

if (!require("ggplot2")) install.packages("ggplot2")
library("ggplot2")

if (!require("ggthemes")) install.packages("ggthemes")
library("ggthemes")
```

# Correspondence 
Please address any questions to Myron Keith Gibert Jr at [mkgibertjr@msn.com](mailto:mkgibertjr@msn.com).  Code for this project is stored in a [GitHub repository](https://github.com/mkgibertjr/CouseraRProgramming_Assignments).

# Introduction
For this assignment, I created two functions.  The first function will create a special list vector that can store an inverted matrix based on an input matrix.  The second function will create an inverted matrix using the special vector and store that matrix within that vector's environment.  In addition to acquiring more practice in writing functions, this assignment involved an understanding of lexical scoping. Because of this, I will take extra steps to explain what each function does, and then perform some tests at the end to confirm that the functions work as intended.

The full instructions for this assignment can be found [here](https://github.com/rdpeng/ProgrammingAssignment2).

## Set Parameters
```{r parameters}
#set output directory? Default: outputdir <- "assignment3outputs"
outputdir <- "assignment3outputs"
#Overwrite contents of the output directory? Default: deleteoutputs <- FALSE
deleteoutputs <- TRUE
#Delete specdata/ directory after completing the analysis? Default: deletespec <- TRUE
deletehos <- TRUE
```

## Debug

The debug chunk will prevent the script from running if any of the dependent variables for this analysis do not exist.  This should prevent the program from erroring out after a long runtime without producing any results due to a missing variable.  If modifying the input .csv and .xlsx files, it is important to leave all header information and column names intact, as the program uses this information to extract relevant data.  Columns are intuitively labeled to end user convenience.

```{r debug}
if (dir.exists(outputdir) && deleteoutputs == FALSE ){
  stop("Your output directory already exists!  Please delete/move 
       this folder from your working directory.  Alternatively, you 
       can set 'deleteoutputs' to TRUE to auto-delete this folder 
       for every run. You may also choose an alternative output 
       directory.")
}else{
  unlink(outputdir,recursive = TRUE)
}

if (!exists("outputdir")){
stop("outputdir variable is not defined.  Please ensure that all 
     parameters in the r parameters chunk are defined.")
}

if (!exists("deleteoutputs")){
stop("deleteoutputs variable is not defined.  Please ensure that all 
     parameters in the r parameters chunk are defined.")
}

if (!exists("deletehos")){
stop("deletehos variable is not defined.  Please ensure that all 
     parameters in the r parameters chunk are defined.")
}

if (!dir.exists(outputdir)){dir.create(outputdir)}
```

# Data

## Unzipping the data

For this programming assignment I needed to unzip this folder. Once I unzipped the zip file, I did not make any modifications to the files in the directories. 

```{r unzip}
if(
  !dir.exists("hospital-data.csv")|
   !dir.exists("outcome-of-care-measures.csv")|
   !dir.exists("Hospital_Revised_Flatfiles.pdf")
){unzip("ASN3_rprog_data_hospital.zip")}

```

# Plot the 30-day mortality rates for heart attack

I first needed to read the outcome data into R via the read.csv function and look at the first few rows.

```{r read30day,results=FALSE,message=FALSE}
outcomedata <- read.csv("outcome-of-care-measures.csv", colClasses = "character")

head(outcomedata)
```

I then need to make a simple histogram of the 30-day death reates from heart attack.  This is column 11 in the outcome dataset.

```{r plot30day}

# I will get a warning about NAs being introduced; that is okay
outcomedata[,11] <- as.numeric(outcomedata[,11])

heartattack <- outcomedata[,11]

heartattack <- as.data.frame(heartattack[complete.cases(heartattack)])

colnames(heartattack) <- "thirty_day_HA_mortality"

p <- ggplot(heartattack,aes(thirty_day_HA_mortality)) + 
  geom_histogram() + 
  xlab("Days") + 
  ylab(expression("Deaths")) + 
  ggtitle(expression("30-Day mortality rates of Heart Attack Patients")) +
  theme_light() +
  scale_color_colorblind() +
  scale_fill_colorblind() +
  theme(axis.text.x=element_text(size=rel(1.2),angle = 90, vjust = 0.5, hjust = 1),
        plot.title = element_text(size=rel(1.2), face="bold",hjust = 0.5),
        legend.position = "none") 

p

```

Because I originally read the data in as character (by specifying colClasses = "character"), I need to coerce the column to be numeric. I got a warning about NAs being introduced but that is okay, as there is missing data.

# Finding the best hospital in a state

```{r findbesthos}

best <- function(state,outcome){
  
  #state <- "TX"
  #outcome <- "heart attack"
  
  data <- read.csv("outcome-of-care-measures.csv", colClasses = "character")
  
  if (is.na(match(state,data$State))){
  stop("invalid state")
  }
  
  if (outcome == "heart attack" | 
      outcome == "Heart Attack" | 
      outcome == "heart failure" | 
      outcome == "Heart Failure" |
      outcome == "pneumonia" |
      outcome == "Pneumonia"){}else{
        stop("invalid outcome")
      }
  
  beststate <- data[which(data$State==state),]
  
  if(outcome=="heart attack" | outcome=="Heart Attack"){
    minrate <- min(
                as.numeric(
                beststate$Hospital.30.Day.Death..Mortality..Rates.from.Heart.Attack)
                ,na.rm=TRUE)
    besthos <- beststate$Hospital.Name[
      match(
        minrate
        ,as.numeric(
        beststate$Hospital.30.Day.Death..Mortality..Rates.from.Heart.Attack))]
  }else{
  if(outcome=="heart failure" | outcome=="Heart Failure"){
    minrate <- min(
                  as.numeric(
                  beststate$Hospital.30.Day.Death..Mortality..Rates.from.Heart.Failure)
                  ,na.rm=TRUE)
    besthos <- beststate$Hospital.Name[
      match(
        minrate
        ,as.numeric(
        beststate$Hospital.30.Day.Death..Mortality..Rates.from.Heart.Failure))]
  }else{
  if(outcome=="pneumonia" | outcome=="Pneumonia"){
    minrate <- min(
                   as.numeric(
                   beststate$Hospital.30.Day.Death..Mortality..Rates.from.Pneumonia)
                   ,na.rm=TRUE)
    besthos <- beststate$Hospital.Name[
      match(
        minrate
        ,as.numeric(
        beststate$Hospital.30.Day.Death..Mortality..Rates.from.Pneumonia))]
  }}}
  
  besthos
  
}

best("TX","Heart Attack")

best("TX","heart failure")

best("MD","heart attack")

best("MD","pneumonia")

#best("BB", "heart attack")

#best("NY","hert attack")

```

```{r findbesthos_run,echo=FALSE}

best <- function(state,outcome){
  
  #state <- "TX"
  #outcome <- "heart attack"
  
  data <- read.csv("outcome-of-care-measures.csv", colClasses = "character")
  
  if (is.na(match(state,data$State))){
  stop("invalid state")
  }
  
  if (outcome == "heart attack" | 
      outcome == "Heart Attack" | 
      outcome == "heart failure" | 
      outcome == "Heart Failure" |
      outcome == "pneumonia" |
      outcome == "Pneumonia"){}else{
        stop("invalid outcome")
      }
  
  beststate <- data[which(data$State==state),]
  
  if(outcome=="heart attack" | outcome=="Heart Attack"){
    minrate <- min(as.numeric(beststate$Hospital.30.Day.Death..Mortality..Rates.from.Heart.Attack),na.rm=TRUE)
    besthos <- beststate$Hospital.Name[match(minrate,as.numeric(beststate$Hospital.30.Day.Death..Mortality..Rates.from.Heart.Attack))]
  }else{
  if(outcome=="heart failure" | outcome=="Heart Failure"){
    minrate <- min(as.numeric(beststate$Hospital.30.Day.Death..Mortality..Rates.from.Heart.Failure),na.rm=TRUE)
    besthos <- beststate$Hospital.Name[match(minrate,as.numeric(beststate$Hospital.30.Day.Death..Mortality..Rates.from.Heart.Failure))]
  }else{
  if(outcome=="pneumonia" | outcome=="Pneumonia"){
    minrate <- min(as.numeric(beststate$Hospital.30.Day.Death..Mortality..Rates.from.Pneumonia),na.rm=TRUE)
    besthos <- beststate$Hospital.Name[match(minrate,as.numeric(beststate$Hospital.30.Day.Death..Mortality..Rates.from.Pneumonia))]
  }}}
  
  besthos
  
}

best("TX","Heart Attack")

best("TX","heart failure")

best("MD","heart attack")

best("MD","pneumonia")

#best("BB", "heart attack")

#best("NY","hert attack")

```

<<<<<<< HEAD
# Ranking hospitals by outcome in a state
=======
## Ranking hospitals by outcome in a state
>>>>>>> cacf3c5edda70402efef67a811aa0a8cd4e32a40

```{r rankhospital}

rankhospital <- function(state,outcome,num = "best"){
  
    
  #state <- "MD"
  #outcome <- "heart attack"
  #num <- "worst"
  
  data <- read.csv("outcome-of-care-measures.csv", colClasses = "character")
  
  if (is.na(match(state,data$State))){
  stop("invalid state")
  }
  
    if (outcome == "heart attack" | 
      outcome == "Heart Attack"){i <- 11} else if(
      outcome == "heart failure" | 
      outcome == "Heart Failure"){i <- 17} else if(
      outcome == "pneumonia" |
      outcome == "Pneumonia"){i <- 23}else{
        stop("invalid outcome")
      }
  
beststate <- data[which(data$State==state),]
beststate[,i] <- as.numeric(beststate[,i])

beststate <- beststate[which(!is.na(beststate[,i])),]

beststate <- beststate[order(beststate[,i], beststate$Hospital.Name), ]

if(num=="best" | num=="Best"){
  num2 <- 1
  }
   
if(num=="worst" | num=="Worst"){
  num2 <- nrow(beststate)  
}else if(is.numeric(num)) {num2 <- num}
  else if(num<1 || num > nrow(beststate)) {
      return(NA)
    } else {
    stop('invalid num')
    }

return.names <- beststate[num2, ]$Hospital.Name

return.names[1]

}

rankhospital("TX","heart failure",4)

rankhospital("MD","heart attack", "worst")

rankhospital("MN","heart attack",5000)

```

<<<<<<< HEAD
# Ranking hospitals in all states
=======
## Ranking hospitals in all states
>>>>>>> cacf3c5edda70402efef67a811aa0a8cd4e32a40

```{r rankall}

rankall <- function(outcome,num = "best"){
  
  #outcome <- "pneumonia"
  #num <- "worst"
  
  data <- read.csv("outcome-of-care-measures.csv", colClasses = "character")
  
    if (outcome == "heart attack" | 
      outcome == "Heart Attack"){i <- 11} else if(
      outcome == "heart failure" | 
      outcome == "Heart Failure"){i <- 17} else if(
      outcome == "pneumonia" |
      outcome == "Pneumonia"){i <- 23}else{
        stop("invalid outcome")
      }
  
data[,i] <- as.numeric(data[,i])

state.i <- 1

states <- unique(as.character(data$State))
ranking <- data.frame(hospital=character(0),state=character(0))

for (state.i in 1:length(states)){
#state <- "WY"
state <- states[state.i]

data.state <- data[which(data$State==state),]

data.state[,i] <- as.numeric(data.state[,i])

data.state <- data.state[which(!is.na(data.state[,i])),]

data.state <- data.state[order(data.state[,i], data.state$Hospital.Name), ]

if(num=="best" | num=="Best"){
  num2 <- 1
  }
   
if(num=="worst" | num=="Worst"){
  num2 <- nrow(data.state)  
}else if(is.numeric(num)) {num2 <- num}
  else if(num<1 || num > nrow(data.state)) {
      return(NA)
    } else {
    stop('invalid num')
  }

return.names <- data.state[num2, ]$Hospital.Name

hospital <- return.names[1]

rbinder <- cbind(state,hospital)

ranking <- rbind(ranking,rbinder)
}
colnames(ranking) <- c("state","hospital")
ranking[order(ranking$state),]
}

test1 <- rankall("heart attack", 20)
head(test1,10)

test2 <- rankall("pneumonia","worst")
tail(test2,3)

```

<<<<<<< HEAD
# Quiz
=======
## Quiz
>>>>>>> cacf3c5edda70402efef67a811aa0a8cd4e32a40

After completing the programming assignment, I had to complete a quiz using the provided code with each question.  I first had to run the provided code, and then select my output from the multiple choice options.  This quiz was used to verify whether my three functions can effectively answer the data science questions from this data set.  My final score was a 100%.

1. What result is returned by the following code?

```{r q1}
best("SC", "heart attack")
```

2. What result is returned by the following code?

```{r q2}
best("NY", "pneumonia")
```

3. What result is returned by the following code?

```{r q3}
best("AK", "pneumonia")
```

4. What result is returned by the following code?

```{r q4}
rankhospital("NC", "heart attack", "worst")
```

5. What result is returned by the following code?

```{r q5}
rankhospital("WA", "heart attack", 7)
```

6. What result is returned by the following code?

```{r q6}
rankhospital("TX", "pneumonia", 10)
```

7. What result is returned by the following code?

```{r q7}
rankhospital("NY", "heart attack", 7)
```

8. What result is returned by the following code?

```{r q8}
r <- rankall("heart attack", 4)
as.character(subset(r, state == "HI")$hospital)
```

9. What result is returned by the following code?

```{r q9}
r <- rankall("pneumonia", "worst")
as.character(subset(r, state == "NJ")$hospital)
```

10. What result is returned by the following code?

```{r q10}
r <- rankall("heart failure", 10)
as.character(subset(r, state == "NV")$hospital)
```

# Cleanup
This final command removes the unzipped "specdata" directory if the deletespec variable is set to TRUE.  This reduces the overall storage burden of this project by removing the files that we no longer need access to. The zipped file remains in the working directory, so "specdata" will be recreated anyways using the command in line 93 (unzip) if it is deleted here.

```{r cleanup}
if(deletehos== TRUE){
file.remove("hospital-data.csv")
file.remove("outcome-of-care-measures.csv")
file.remove("Hospital_Revised_Flatfiles.pdf")
}
```

