---
title: 'R Programming - Assignment #1'
author: "Myron Keith Gibert Jr"
date: "January 21, 2020"
output: pdf_document
---
##Introduction
For this first programming assignment I wrote three functions that are meant to interact with dataset that accompanies this assignment. The dataset is contained in a zip file specdata.zip that is included in the GitHub repository.

# Data
The zip file containing the data can be downloaded here:

[specdata.zip 2.4MB](https://d396qusza40orc.cloudfront.net/rprog%2Fdata%2Fspecdata.zip)

The zip file contains 332 comma-separated-value (CSV) files containing pollution monitoring data for fine particulate matter (PM) air pollution at 332 locations in the United States. Each file contains data from a single monitor and the ID number for each monitor is contained in the file name. For example, data for monitor 200 is contained in the file "200.csv". Each file contains three variables:

* Date: the date of the observation in YYYY-MM-DD format (year-month-day)
* sulfate: the level of sulfate PM in the air on that date (measured in micrograms per cubic meter)
* nitrate: the level of nitrate PM in the air on that date (measured in micrograms per cubic meter)



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require("tidyverse")) install.packages("tidyverse")
library("tidyverse")
```

## Unzipping the data

For this programming assignment I needed to unzip this file and create the directory 'specdata'. Once I unzipped the zip file, I did not make any modifications to the files in the 'specdata' directory. In each file you'll notice that there are many days where either sulfate or nitrate (or both) are missing (coded as NA). This is common with air pollution monitoring data in the United States.

```{r unzip}
if(!dir.exists("specdata")){
  unzip("rprog_data_specdata.zip")
}
#setwd("./specdata")
```

## Part 1

Write a function named 'pollutantmean' that calculates the mean of a pollutant (sulfate or nitrate) across a specified list of monitors. The function 'pollutantmean' takes three arguments: 'directory', 'pollutant', and 'id'. Given a vector monitor ID numbers, 'pollutantmean' reads that monitors' particulate matter data from the directory specified in the 'directory' argument and returns the mean of the pollutant across all of the monitors, ignoring any missing values coded as NA. A prototype of the function is as follows


```{r pollutantmean}
pollutantmean <- function(directory, pollutant, id){
  #id <- 332
  #directory <- "./specdata"
  #pollutant <- "sulfate"
  filenames <- list.files(directory)
  data <- data.frame()
  for(i in 1:id){
    data <- rbind(data,read.csv(paste(directory,filenames[i],sep = "/")))
  }
  
  mean_pollutant <- mean(data[,pollutant],na.rm = TRUE)
  write.csv(mean_pollutant,file = "mean_pollutant.csv")
}

pollutantmean("./specdata","sulfate",332)

pollutantmean("./specdata","nitrate",332)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
